---
title: "Solving problems with zusmt"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Solving problems with zusmt}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

An SMT solver answers one question: is there an assignment of values to these variables that makes
all of these statements true at once? It differs from an optimiser in that it looks for *any*
solution rather than the best one, and from a constraint solver in that the statements may involve
several theories — integers, reals, uninterpreted functions, arrays — reasoned about together.

```{r setup}
library(zusmt)
```

## Asking a question

A solver is created for a logic, given assertions, and then asked to check them.

```{r}
s <- smt_solver("QF_LIA")
smt_assert(s, "
  (declare-const x Int)
  (declare-const y Int)
  (assert (> x 3))
  (assert (< x 7))
  (assert (= y (* 2 x)))
")

smt_check(s)
```

`"sat"` means a solution exists, and `smt_model()` produces one.

```{r}
smt_model(s)
```

The answer is *a* solution, not *the* solution: any `x` strictly between 3 and 7 would do. If you
need a particular one, constrain it further.

## When there is no solution

```{r}
u <- smt_solver("QF_LIA")
smt_assert(u, "(declare-const x Int) (assert (= (* 2 x) 3))")
smt_check(u)
```

`2x = 3` has no integer solution. Note that the same assertion is satisfiable over the reals, which
is the difference between `QF_LIA` and `QF_LRA` rather than a detail of how the problem is written:

```{r}
r <- smt_solver("QF_LRA")
smt_assert(r, "(declare-const x Real) (assert (= (* 2.0 x) 3.0))")
smt_check(r)
smt_model(r)$x
```

Asking for a model when the answer was `"unsat"` is an error rather than an empty list, because
there is nothing to report:

```{r, error = TRUE}
smt_model(u)
```

## Why there is no solution

`"unsat"` on its own does not say which assertions are to blame. A solver created with
`unsat_cores = TRUE` can report the subset that already conflicts, which is usually a small part of
a large problem:

```{r}
conflict <- smt_solver("QF_LIA", unsat_cores = TRUE)
smt_assert(conflict, "
  (declare-const x Int)
  (declare-const y Int)
  (assert (! (> x 10) :named x-large))
  (assert (! (< x 5)  :named x-small))
  (assert (! (= y 1)  :named y-fixed))
")
smt_check(conflict)
smt_unsat_core(conflict)
```

`y-fixed` is absent because it is not part of the contradiction -- that is the point of a core. Name
assertions with `(! ... :named n)` and the names come back with them.

The argument goes on `smt_solver()` rather than being set later because OpenSMT decides whether to
record a proof when it builds the solver. Asking afterwards cannot work, and is refused:

```{r, error = TRUE}
smt_assert(smt_solver("QF_LIA"), "(set-option :produce-unsat-cores true)")
```

## Exact values

Solvers compute in rationals. Many rationals are not doubles, so a model value carries both: the
`numeric` you can compute with, and the solver's exact answer as an attribute.

```{r}
third <- smt_solver("QF_LRA")
smt_assert(third, "(declare-const v Real) (assert (= (* 3.0 v) 1.0))")
smt_check(third)

value <- smt_model(third)$v
value
attr(value, "exact")
```

Whether that distinction matters depends on what you do next. It matters if you feed the answer back
into another constraint.

## Beyond arithmetic

Uninterpreted functions let you state that something *is a function* — equal inputs give equal
outputs — without saying which function. That alone is enough to make some problems unsatisfiable:

```{r}
uf <- smt_solver("QF_UF")
smt_assert(uf, "
  (declare-sort U 0)
  (declare-fun f (U) U)
  (declare-const a U)
  (declare-const b U)
  (assert (= a b))
  (assert (distinct (f a) (f b)))
")

smt_check(uf)
```

No definition of `f` can satisfy that, whatever `U` and `f` are.

Arrays work similarly, with `select` and `store` and the rule that reading back what you just wrote
gives what you wrote:

```{r}
ax <- smt_solver("QF_AX")
smt_assert(ax, "
  (declare-sort I 0)
  (declare-sort E 0)
  (declare-const arr (Array I E))
  (declare-const i I)
  (declare-const e E)
  (assert (distinct (select (store arr i e) i) e))
")

smt_check(ax)
```

## Building up a problem

Assertions accumulate, so a problem can be built across several calls — and `push`/`pop` let you
explore an assumption and then discard it.

```{r}
k <- smt_solver("QF_LIA")
smt_assert(k, "(declare-const n Int) (assert (> n 10))")
smt_check(k)

smt_assert(k, "(push 1) (assert (< n 5))")
smt_check(k)

smt_assert(k, "(pop 1)")
smt_check(k)
```

`smt_assert()` accepts any SMT-LIB2 command, not only `assert`, so a script written for another
solver usually runs unchanged.

## Which logic to ask for

```{r}
smt_logics()
```

Pick the weakest logic that expresses the problem: the solver can use a specialised decision
procedure, and a mistake in the problem is more likely to be reported as an error than quietly
accepted. `QF_IDL` and `QF_RDL` are difference logic — constraints of the form `x - y <= c` — which
covers scheduling and similar problems and is decided faster than general arithmetic.

## Errors

Mistakes in the input become R conditions rather than printed output, so they can be caught:

```{r, error = TRUE}
bad <- smt_solver("QF_LIA")
smt_assert(bad, "(assert (> undeclared 1))")
```

## Releasing a solver

A solver holds memory in the bundled C++ library, freed when R garbage-collects the handle. To
release it at a known point — a long loop, a large problem — use `smt_release()`.

```{r}
smt_release(s)
```
